---
title: "Data Analysis"
author: "Juste Simanauskaite & Patricia Rivera"
output: 
  pdf_document: 
    highlight: zenburn
    toc: true
    toc_depth: 6
    
  html_document: default

---

```{r global_options, include=TRUE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE, fig.height=3, fig.width=5, fig.align="center")
library(tidyverse)
library(broom)
library(plyr)
library(survival)
library(survminer)
aids <- read.csv( "http://pages.pomona.edu/~jsh04747/courses/math150/AIDSdata.csv")
dim(aids)
summary(aids)

```

The data set contains a sample size equal to 851 participants and 16 different variables.

```{r}
library(plotrix)
male<-sum(aids$sex==1)
female<-sum(aids$sex==2)
slices <- c(male, female) 
lbls <- c("Male", "Female")
pct <- round(slices/sum(slices)*100)
lbls <- paste(lbls, pct) 
lbls <- paste(lbls,"%",sep="") 
pie3D(slices,labels=lbls,explode=0.1,
   main="Gender Distribution ", cex.lab=0.1)
```
The Pie Chart represents the gender distribution in the sample, with 84% male and 16% female. This shows the potential for the data to not be able to correctly represent the difference of the data variance by gender, if there were to be one. Therefore, gender is something to look into in future data analysis.
```{r}
wnh<-sum(aids$raceth==1)
bnh<-sum(aids$raceth==2)
h<-sum(aids$raceth==3)
api<-sum(aids$raceth==4)
aian<-sum(aids$raceth==5)
oth<-sum(aids$raceth==6)
slices <- c(wnh,bnh,h,api,aian,oth) 
lbls <- c("White Non-Hispanic", "Black Non-Hispanic", "Hispanic","Asian, Pacific Islander", "American Indian, Alaskan Native", "Other/unknown")
pct <- round(slices/sum(slices)*100)
lbls <- paste(lbls, pct) 
lbls <- paste(lbls,"%",sep="") 
pie(slices,lbls,col = rainbow(length(lbls)), cex=0.5 )

```
The distribution od race/ethnicity shows that the greatest number of participants consists of white non-hispanic identifying indiciduals, with black non-hispanic following and hispanic as the 3rd largest represented group.
```{r}
never<-sum(aids$ivdrug==1)
cur<-sum(aids$ivdrug==2)
prev<-sum(aids$ivdrug==3)
slices <- c(never,cur,prev) 
lbls <- c("Never", "Currently", "Previously")
pct <- round(slices/sum(slices)*100)
lbls <- paste(lbls, pct) 
lbls <- paste(lbls,"%",sep="") 
pie3D(slices,labels=lbls,explode=0.1,col=c("turquoise","magenta","salmon"),cex.sub=0.5,
   main="IV Drug Use History ")

```
From this chart we see that most of the participants (84%) have never used IV drugs, whereas 16% of participants have some type of history of usage and none of the participants reported to be currently using the drugs.

```{r}
hist(aids$time)

###Data Plots

fit <- survfit(Surv(time,censor)~tx, data = aids)
ggsurvplot(fit,data = aids,conf.int = FALSE)

aids_fit_time <- survfit(Surv(time, censor) ~ sex, data=aids)
ggsurvplot(aids_fit_time, data=aids,  conf.int = TRUE)

aids_fit_time.d <- survfit(Surv(time_d, censor_d) ~ sex, data=aids)
ggsurvplot(aids_fit_time.d, data=aids,  conf.int = TRUE)

```
#Survival Analysis 
```{r}
#mutation of age
aids <- read.csv( "http://pages.pomona.edu/~jsh04747/courses/math150/AIDSdata.csv")
aids <- aids %>% 
  mutate(age = ifelse(age <= 20, "under20", 
                             ifelse(age <=30, "20-30",
                                    ifelse(age <= 40, "30-40",
                                           ifelse(age <=50, "40-50",
                                                 ifelse(age <=60, "50-60",
                                                        ifelse(age <=70, "60-70", "over70"))))))) %>%
  mutate(age = factor(age,
                       levels = c("under20", "20-30", "30-40","40-50", "50-60","60-70","over70")),  sex = ifelse(sex == 2, "male","female"))
```

```{r}
library(survival)
library (survminer)
library(ggplot2)
library(broom)

coxph(Surv(time_d,censor_d) ~ sex , data=aids)  %>% tidy()
coxph(Surv(time,censor) ~ sex, data=aids) %>% tidy()

coxph(Surv(time,censor) ~ age+ txgrp+ karnof, data=aids) %>% tidy()
cox.zph(coxph(Surv(time,censor) ~ age + txgrp+karnof, data=aids))
coxph(Surv(time,censor) ~ age *txgrp*karnof, data=aids) %>% tidy()
cox.zph(coxph(Surv(time,censor) ~ age *txgrp*karnof, data=aids))

ggsurvplot(survfit(Surv(time,censor) ~ 1, data=aids), 
           censor=F, conf.int=T, fun="cumhaz") + ggtitle("Estimated Hazard rates")


ggsurvplot(survfit(Surv(time,censor) ~ sex, data=aids), 
           censor=F, conf.int=T, fun="cumhaz") + ggtitle("Estimated Hazard rates based on sex")
ggsurvplot(survfit(Surv(time,censor) ~ txgrp, data=aids), 
           censor=F, conf.int=T, fun="cumhaz") + ggtitle("Estimated Hazard rates based on treatment group")

ggsurvplot(survfit(Surv(time,censor) ~ karnof, data=aids), 
           censor=F, conf.int=T, fun="cumhaz") + ggtitle("Estimated Hazard rates based on klarnfsky")

ggsurvplot(survfit(Surv(time,censor) ~ age, data=aids), 
           censor=F, conf.int=T, fun="cumhaz") + ggtitle("Estimated Hazard rates based on age")





#ggsurvplot(survfit(Surv(time, censor)~hemophil, data = aids),
           #censor=F, conf.int = T, fun = "cumhaz")


ggsurvplot(survfit(Surv(time, censor)~ivdrug, data = aids),
           censor=F, conf.int = T, fun = "cumhaz")


coxph(Surv(time,censor) ~ ivdrug, data=aids) %>% tidy()

coxph(Surv(time,censor) ~ ivdrug*karnof, data=aids) %>% tidy()
#how tow modify so that sex is labeled as male and female



coxph(Surv(time,censor)~sex+tx+age+txgrp, data = aids) %>% tidy()

```



#Patricia's "Something New"
I will be doing a power analysis by simulating survival analysis curves

### 1. What is the topic?
The topic is using sim.survdata in R to simulate survival data. Using that simulated data, we will make that the alternative and control for the coefficient beta by setting it equal to some value. Then using power analysis, we will see how many times we reject $H_0$. 

### 2. How it is relevant? How it relates to survival analysis/analysis at hand?
Power analysis relates to survival analysis because if power is large after comparing our data to the simulated survival data, this tells us that there is a high chance that we would reject the null in favor of the alternative (control versus treatment?)

### 3. Resources to learn about the topic.
Below are some of the resources I have begun to use to learn about creating simulations of survival curves and performing power analysis:

a). https://cran.r-project.org/web/packages/coxed/vignettes/simulating_survival_data.html
b). http://www.icssc.org/documents/advbiosgoa/tab%2026.00_survss.pdf

### 4. What will be challenging about learning something new?
Learning something new will be challenging because in this case, the concept of power analysis is something I just recently learned in Intro to Statistics. So learning to apply this concept in the context of survival analysis curves will be a challenge for me to learn. Learning how to simulate survival curves will also be challenging because I will have to learn how to use and interpret new functions in R.
=======
# Juste's "Something New"
I will be analyzing the Shoenfeld residuals for the Cox PH model.

### 1. What is goign on? What is the topic? 2. How it is relevant? How it relates to survival analysis/analysis at hand?

Cox proportional hazards (PH) model is considered a great way to identify combined effects of several covariates on the relative risk (hazard). This model assumes that the hazards of the different strata formed by the levels of the covariates are proportional. This proportional hazards assumption is particularly important and can be tested via three different clases of tests. The first class is focused on the piecewise estimation of models for subsets of data defined by stratification of time. The second one considers the interactions between covariates and some function of time. Final, third one is based on examinations of regression residuals. The Schoenfeld Residuals are a part of the third class of proportional hazard assumption testing and I will be exploring it in order to be able to eradicate a method for testing for the PH assumption in the current and future data set analyses. This topic is particularly important in relation to survival analysis since it provides an idea of whether the model is appropriate for the data set at hand and whether some covariates should be considered as variants of time in order to supply the best model for prediction of proportional hazards.

### 3. Resources to learn about the topic.

I have been researching articles and scientific journals that provide insights into this model and comparisons between the Cox PH and teh parametric model. Sources include: 
a) https://krex.k-state.edu/dspace/bitstream/handle/2097/8787/AngelaCrumer2011.pdf 
b) http://nematilab.info/bmijc/assets/weibull_cox.pdf 
c) https://www.jstatsoft.org/article/view/v070i08 

### 4. What will be challenging about learning something new?

Taking a completely new model of analyzing survival data is particulalrly difficult since the mathematical derivations and notations are also very varied from what we have seen in class. Although, I do remember some of the ideas behind parametric functions, their applications to statistical models are much more challenging than I have expected. Therefore, it will require me a lot of time and extensive research to be able to understand and learn how to apply this model to our data and other instances of survival analysis.

```{r}
### some trials of applications of parametric functions in r
library(flexsurv)

flexsurvreg(Surv(time, censor) ~ age, data = aids,  dist = "weibull")

```
#More about the Weibull Model
The Weibull model is very similar to the Cox PH model we have explored in class. The Weibull Model usually used when the exponential distribution is not suffiecient to come up with a model. 
The exponential desnity function is 
$f(t)= \lambda exp(-\lambda(t))$, for $\lambda>0$ and $t>0$
With a constant hazard function of $h(t)=\lambda$

```{r}
letters

```

SHoenfeld:
```{r}
veteran.ph <- coxph(Surv(time,status) ~ trt, data=veteran) 
cox.veteran <- cox.zph(veteran.ph)
cox.veteran
ggcoxzph(cox.veteran)
ggcoxdiagnostics(veteran.ph, type="schoenfeld")
```


